---
title: "Data Visualization Using R & ggplot2"
author: "Naupaka Zimmerman (@naupakaz) and Andrew Tredennick (@ATredennick) (and Karthik Ram (@_inundata))"
date: "August 6, 2017"
header-includes:
   - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\},fontsize=\footnotesize}
   - \makeatletter
   - \def\verbatim{\footnotesize\@verbatim \frenchspacing\@vobeyspaces \@xverbatim}
   - \makeatother
output:
  beamer_presentation:
    theme: "metropolis"
    colortheme: "default"
    fig_caption: no
    slide_level: 2
    keep_tex: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=2.5, out.width='.75\\linewidth', fig.align='center')
```

## Some housekeeping
Install some packages
```{r house, echo=T, eval=F}
install.packages("ggplot2", dependencies = TRUE)
install.packages("ggthemes")
install.packages("tidyr")
install.packages("dplyr")
```

```{r house2, echo=F, eval=T, message=F}
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)
```

# Why ggplot?

## 1 minute
```{r gap, echo=F, eval=T, fig.width=5, fig.height=4}
library(ggplot2)
library(gapminder)
library(ggthemes)

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(shape = 1, aes(color = continent)) +
  stat_smooth(method = "lm", size = 1, color = "black") +
  scale_x_log10() + 
  xlab("Per Capita GDP") + 
  ylab("Life Expectancy (yrs)") +
  facet_wrap(~continent) +
  theme_few() + 
  guides(color = FALSE)
```

## 1 minute
```{r gap2, echo=T, eval=F, fig.width=5, fig.height=4}
library(ggplot2)
library(gapminder)
library(ggthemes)

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(shape = 1, aes(color = continent)) +
  stat_smooth(method = "lm", size = 1, color = "black") +
  scale_x_log10() + 
  xlab("Per Capita GDP") + 
  ylab("Life Expectancy (yrs)") +
  facet_wrap(~continent) +
  theme_few() + 
  guides(color = FALSE)
```

## 30 minutes
```{r gap3, echo=FALSE, eval=TRUE,  fig.width=5, fig.height=4}
library(scales)
library(gapminder)
gapminder <- as.data.frame(gapminder)
conts <- unique(gapminder[,"continent"])
cols <- scales::hue_pal()(length(conts))
par(mfrow = c(2,3))
counter <- 1
for (i in conts) {
  plot(gapminder[which(gapminder$continent == i), "gdpPercap"],
       gapminder[which(gapminder$continent == i), "lifeExp"], col = cols[counter],
       xlab = "Per Capita GDP", ylab = "Life Expectancy (yrs)",
       main = i, las = 1, log = "x")
  fit <- lm(gapminder[which(gapminder$continent == i), "lifeExp"] ~ log(gapminder[which(gapminder$continent == i), "gdpPercap"]))
  pred <- predict(fit, interval = "confidence")
  lines(sort(gapminder[which(gapminder$continent == i), "gdpPercap"]), sort(pred[,1]))
  lines(sort(gapminder[which(gapminder$continent == i), "gdpPercap"]), sort(pred[,2]), lty = 2)
  lines(sort(gapminder[which(gapminder$continent == i), "gdpPercap"]), sort(pred[,3]), lty = 2)
  counter <- counter + 1
}
```

## 30 minutes
```{r gap4, echo=T, eval=F}
library(scales)
library(gapminder)
gapminder <- as.data.frame(gapminder)
conts <- unique(gapminder[,"continent"])
cols <- scales::hue_pal()(length(conts))
par(mfrow = c(2,3))
counter <- 1
for (i in conts) {
  plot(gapminder[which(gapminder$continent == i), "gdpPercap"],
       gapminder[which(gapminder$continent == i), "lifeExp"], col = cols[counter],
       xlab = "Per Capita GDP", ylab = "Life Expectancy (yrs)",
       main = i, las = 1, log = "x")
  fit <- lm(gapminder[which(gapminder$continent == i), "lifeExp"] ~ 
```

## But wait, there's more...
```{r gap5, echo=T, eval=F}
log(gapminder[which(gapminder$continent == i), "gdpPercap"]))
  pred <- predict(fit, interval = "confidence")
  lines(sort(gapminder[which(gapminder$continent == i), "gdpPercap"]), sort(pred[,1]))
  lines(sort(gapminder[which(gapminder$continent == i), "gdpPercap"]), sort(pred[,2]), lty = 2)
  lines(sort(gapminder[which(gapminder$continent == i), "gdpPercap"]), sort(pred[,3]), lty = 2)
  counter <- counter + 1
}
```


# Why ggplot?

##  Why ggplot?
* More elegant and compact code than with base graphics
* More aesthetically pleasing defaults than lattice
* Very powerful for exploratory data analysis

## Why ggplot?
* \alert{gg} is for \alert{grammar of graphics} (term by Lee Wilkinson)
* A set of terms that defines the basic components of a plot
* Used to produce figures using coherant, consistant syntax

##  Why ggplot2?
* Supports a continuum of expertise
* Easy to get started, plenty of power for complex figures 


# The Grammar

##  Some terminology
### Data
* Must be a data frame (```data.frame()```, ```as.data.frame()```)
* Gets pulled into the ```ggplot()``` object

##  tidyr
Helps your data play nice with ggplot


# Aesthetics

## Some terminology
### \alert{aes}thetics
* How your data are represented visually 
  + a.k.a. mapping
* which data on the x
* which data on the y
* but also: \textcolor{red}{color}, \LARGE{size}, \normalsize{shape, transparency}

# Geometries

## Some terminology
### \alert{geom}etry
* The geometric objects in the plot
* points, lines, polygons, etc.
* functions: ```geom_point()```, ```geom_bar()```, ```geom_line()```

##  Basic structure
```{r strct-basic, echo=T, eval=F}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

# Equivalently...
myplot <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
myplot + geom_point()
```

* Specify the data and variables inside the ```ggplot()``` function.
* Anything else that goes in here becomes a \emph{global} setting.
* Then add layers: geometric objects, statistical models, and facets.

##  Don't be tempted!
* \alert{NEVER} use ```qplot()``` - short for quick plot.
* You‘ll end up unlearning and relearning a good bit.

# Exercise 1
##  Exercise 1
```{r d2-gen, eval=T, echo=T}
# Make a small sample of the diamonds dataset
d2 <- diamonds[sample(x = 1:nrow(diamonds), size = 1000), ]
```
The make this plot:
```{r d2-plot, eval=T, echo=F}
ggplot(data = d2, aes(x = carat, y = price, color = color)) +
  geom_point()
```

# Stats
##  Some terminology
### \alert{stats}
* Statistical transformations and data summary
* All geoms have associated default stats, and vice versa
* e.g. binning (the stat) for a histogram or fitting a linear model


# Facets
##  Some terminology
### \alert{facet}s
* Subsetting data to make lattice plots
* Really powerful
* I use in almost every publication

# Scales
## Some terminology
### scales
* Control the \emph{mapping} from data to aesthetics
* Often used for adjusting color mapping (i.e., setting colors manually)

##  Refer to a color chart for beautiful visualizations

\begin{center}
\url{http://tools.medialab.sciences-po.fr/iwanthue/}
\newline
\newline
\includegraphics[scale=0.25]{images/color_schemes.png}
\end{center}

## Some common scales
```{r com-scales, eval=F}
scale_fill_discrete(); scale_colour_discrete()
scale_fill_hue(); scale_color_hue()
scale_fill_manual();  scale_color_manual()
scale_fill_brewer(); scale_color_brewer()
scale_linetype(); scale_shape_manual()
```

# Coordinates
## Some terminology
### \alert{coord}inates
* Not going to cover this in detail
* e.g. polar coordinate plots

##  Don't try this at home...please!
```{r makdat, eval=T, echo=F}
mydata <- data.frame(group = c(LETTERS[1:4]),
                     proportion = c(0.25,0.50,0.20,0.05))
```

```{r pie, eval=T, echo=T}
ggplot(mydata, aes(x = "", y = proportion, fill = group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start=0) +
  theme_void()
```

# Putting it all together with more examples
# Histograms

# Line plots

# Exercise 2
##  Exercise 2
* Modify the last plot and change the ribbon to three separate lines, like below:
```{r conf-ex, echo=F}
climate <- read.csv("../data/climate.csv", header = TRUE)
ggplot(climate, aes(Year, Anomaly10y)) +
  geom_line(aes(y = Anomaly10y - Unc10y, x = Year), color="red", linetype=2) +
  geom_line(aes(y = Anomaly10y + Unc10y, x = Year), color="red", linetype=2) +
  geom_line(color = "black")
```

# Bar plots

# Exercises 3, 4, & 5
##  Exercise 3
Using the ```d2``` dataset you created earlier, make this plot:
```{r d2-bar, echo=F}
ggplot(d2, aes(x = clarity, fill = cut)) +
  geom_bar(position = "dodge")
```

##  Exercise 4
Use ```dplyr``` to summarize the iris data, then make the plot below.
```{r iris-summ}
iris_agg <- iris %>%
  group_by(Species) %>%
  summarise(mean_petal_len = mean(Petal.Length),
            sd_petal_len = sd(Petal.Length))
# ?geom_errorbar()
```

```{r iris-summ2, echo=F}
ggplot(iris_agg, aes(x = Species, y = mean_petal_len, color = Species)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean_petal_len - sd_petal_len,
                    ymax = mean_petal_len + sd_petal_len), width = 0.15) +
  xlab("Species") +
  ylab("Mean Petal Length") +
  guides(color=FALSE) +
  theme_few()
```

##  Exercise 5
* Using the ```climate``` dataset, create a new variable called ```sign```. Make it categorical (above/below) based on the sign of ```Anomaly10y```.
* Plot a bar plot and use sign variable as the fill.
```{r climanom, echo=F}
climate$sign <- "above"
climate[which(climate$Anomaly10y < 0), "sign"] <- "below"
ggplot(climate, aes(x = Year, y = Anomaly10y, fill = sign)) +
  geom_col() +
  scale_fill_manual(values = c("dodgerblue", "salmon"))
```

# Density plots

# Adding smoothers

# Themes!
##  Adding themes
* Everything can be customized using ```theme()``` settings
* http://ggplot2.tidyverse.org/reference/theme.html

##  What about Wes Anderson?
\begin{center}
\includegraphics[scale=.12]{images/tenenbaums.png}
\end{center}

##  There's a theme for that!
```{r wes}
# install.packages('wesanderson')
library("wesanderson") 
# display a palette 
wes_palette("Royal1")
```

# Save your beautiful plot
##  ggsave
* If the plot is on the screen
```{r save1, eval=F}
ggsave('~/path/to/figure/filename.png')
```

* If your plot is assigned to a named object
```{r save2, eval=F}
ggsave(plot1, '~/path/to/figure/filename.png')
```

* Specify size
```{r save3, eval=F}
ggsave(file = '~/path/filename.png', width = 4, height = 3, units = "in")
```

* And format
```{r save4, eval=F}
ggsave(file = "/path/to/figure/filename.eps")
ggsave(file = "/path/to/figure/filename.jpg")
ggsave(file = "/path/to/figure/filename.pdf")
```

##  Further help
* We've only scratched the surface
* Practice
* Read: http://ggplot2.tidyverse.org/index.html
* Work together

\begin{center}
\includegraphics[scale=.15]{images/chang_book.png}
\includegraphics[scale=.15]{images/hadley.png}
\end{center}
